---
title: "Exam 3 - Connor Dority"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(broom)
library(tidyr)
library(purrr)
```


# 1 — FACULTY DATA

## Read and clean data
```{r}
faculty_raw <- read_csv("FacultySalaries_1995.csv") |>
  clean_names()

faculty <- faculty_raw |>
  select(
    state, tier,
    avg_assist_prof_salary,
    avg_assoc_prof_salary,
    avg_full_prof_salary
  ) |>
  pivot_longer(
    cols = starts_with("avg_"),
    names_to = "rank",
    values_to = "salary"
  ) |>
  mutate(
    rank = case_when(
      rank == "avg_assist_prof_salary" ~ "Assist",
      rank == "avg_assoc_prof_salary" ~ "Assoc",
      rank == "avg_full_prof_salary" ~ "Full"
    )
  )

glimpse(faculty)

p1 <- ggplot(faculty, aes(x = rank, y = salary, fill = rank)) +
  geom_boxplot() +
  facet_wrap(~ tier) +
  labs(x = "Rank", y = "Salary") +
  theme_bw()

p1

```

## ANOVA model

```{r}
anova_model <- aov(salary ~ state + tier + rank, data = faculty)
summary(anova_model)
```

## Interpretation

Salary differs significantly by rank, with Full professors earning the highest salaries, followed by Associate and Assistant professors. Additionally, institutional tier influences salary, with Tier I schools generally showing higher salary distributions compared to Tier II and IIB institutions. This suggests both rank and institutional prestige contribute to salary differences.


# 2 — JUNIPER DATA


## Read and clean data

```{r}
juniper <- read_csv("Juniper_Oils.csv") |>
  clean_names()

glimpse(juniper)
```

## Convert to tidy format

```{r}
jun_long <- juniper |>
  pivot_longer(
    cols = c(
      alpha_acorenol, alpha_cedrene, alpha_chamigrene, alpha_eudesmol,
      alpha_himachalene, alpha_pinene, alpha_terpineol, beta_acorenol,
      beta_cedrene, beta_chamigrene, beta_eudesmol, cedr_8_en_13_ol,
      cedr_8_en_15_ol, cedr_9_ene, cedrol, cis_thujopsene,
      compound_1, compound_2, cuparene, gamma_eudesmol,
      para_cymene, thujopsenal, widdrol
    ),
    names_to = "chemical",
    values_to = "concentration"
  )

glimpse(jun_long)
```

## Plot trends over time

```{r}
p2 <- ggplot(jun_long, aes(x = years_since_burn, y = concentration)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = TRUE) +
  facet_wrap(~ chemical, scales = "free_y") +
  theme_bw()

p2
```

## Build Generalized Linear Model

```{r}
glm_model <- glm(concentration ~ chemical * years_since_burn, data = jun_long)

glm_results <- tidy(glm_model)

glm_results
```

## Find significant relationships

```{r}
significant <- glm_results |>
  filter(str_detect(term, "years_since_burn"), p.value < 0.05)

significant
```

## Interpretation

Several chemical compounds show statistically significant relationships with years since burn, suggesting that fire history influences chemical concentration patterns over time.

